NMRlipids Databank: Making data-driven analyses of membrane properties accessible for all

### 2022-12-06 main.tex
Update main.tex

%\subsection{Finding the best models for PC and PE lipid mixtures}
The power of the NMRlipids databank in selecting the best simulation parameters for a specific application is demonstrated for mixtures of PE and PC lipids in Fig.~\ref{fig:POPCPOPEapls}, showing areas per lipid from available POPC and POPE bilayer simulations in the databank from different force fields at 310\,K. Predictions from different force fields deviate significantly in terms of absolute values and the slope of decrease upon addition of POPE. Because the area per lipid correlates with the average order parameter of the $\it{sn}$-1 chain (Fig.~\ref{fig:quality} F), simulations with the best predictions for area per lipids can be selected based on order parameter quality evaluation. Quality evaluation in Fig.~\ref{fig:quality} and direct comparison in Fig.~\ref{fig:POPC_POPE_dataSI} reveal that Slipids ranks 1st for POPC with a clear advantage to the others, and 2nd for POPE with only marginally lower quality score than GROMOS-CKP and is therefore the best avilable choice for studies where packing effects of PE lipids are important. 
%The direct comparison with experiments in Fig.~\ref{fig:POPC_POPE_dataSI} in the supplementary information reveals that 
Simulations with CHARMM36 and GROMOS-CKP predict too packed bilayer for POPC (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}), while area per lipid for POPC is overstimated in Lipid17. For POPE, 
%order parameters in  Slipids and GROMOS-CKP are close to experiments suggesting that their are per lipids are most reasonable, while
CHARMM36 and Lipid17 predict too packed membranes.

In conclusion, 
%the quality evaluation based on the NMRlipids databank suggests that the Slipids parameters would give the most reliable description for membrane packing of POPC:POPE mixtures. Slipids was a relatively good model also for POPC:POPS mixtures in similar comparisons utilizing the preliminary version of the NMRlipids databank~\cite{antila22b}, although charged membranes are more complicated as the counterion binding affinity plays a significant role in membrane packing~\cite{antila19}. On the other hand, CHARMM36 overcomes Slipids when the focus is on conformational ensembles of headgroups  in different types of lipids because the accuracy of glycerol backbone is low in Slipids~\cite{bacle21}. Altogether, these examples demonstrate how 
the automatic quality evaluation in the NMRlipids databank enables rapid selection of the best simulation parameters to specific applications without extensive and tedious manual force field evaluation. This possibility will promote more reliable simulation results for wide range of applications, support force field development and parametrization of coarse grained force fields against the best atomistic MD simulations. The automatically extracted quantitative quality measures will be particularly useful to guide automatic parametrization procedures.

@@

\subsection{Detection rare phenomena using NMRlipids databank: Cholesterol flip-flops}
Lipid flip-flops from one bilayer leaflet to another play an important role in lipid trafficking and regulating membrane properties~\cite{vanmeer08}. Phospholipid flip-flops are slow, occuring with the timescales of hours or days, while cholesterol, diacylglycerol and ceramide flip-flops are faster, yet the reported timescales range between minutes to sub-millisecods~\cite{vanmeer08,steck12,parisio16,gu19}. These timescales were previously accessible only by coarse-grained simulations or free energy calculations~\cite{parisio16}, yet atomistic simulations reporting cholesterol flip-flop events have been published recently~\cite{gu19,javanainen19,baral20}. 
%However, systematic studies on how cholesterol flip-flop rate depends on membrane properties are still beyond the current computational resources available for a single research group. 
These studies report increase in cholesterol flip-flop rates with increasing unsaturation level and decreasing cholesterol concentration~\cite{gu19,javanainen19}, but correlations between cholesterol flip-flop rates and membrane properties have not been systematically studied. Here we demonstrate that the large amount of simulation data available in the NMRlipids databank with programmatic access makes the analysis of these correlations accessible for all.

Flip-flops were observed for cholesterol, % (634 events in 77 different simulations), 
DCHOL (18,19-di-nor-cholesterol), %, 16 flip-flops in 1 simulation) 
DOG (1,2-dioleoyl-sn-glycerol), %, 44 flip-flops in 4 simulations) 
and SDG (1-stearoyl-2-docosahexaenoyl-sn-glycerol) in the NMRlipids databank. Cholesterol flip-flop rates range between 0.001-1.6\,$\mu$s$^{-1}$ with the mean of 0.16\,$\mu$s$^{-1}$ and median of 0.07\,$\mu$s$^{-1}$. These values are in line with the previously reported values from atomistic MD simulations~\cite{gu19,javanainen19,baral20}. Flip-flop rates of DCHOL, 0.2\,$\mu$s$^{-1}$, was close to the average value of cholesterol, while average rates for diacylglycerols DOG and SDG were higher than for cholesterol, 0.4\,$\mu$s$^{-1}$ and 0.5\,$\mu$s$^{-1}$, respectively. Flip-flops were not observed for other lipids, giving the upper limits for PC lipid flip-flop rate of 9\,s$^{-1}$ and for ceramide (N-palmitoyl-D-erythro-sphingosine) of 0.002\,$\mu$s$^{-1}$. Therefore, data in the NMRlipids databank suggests that the lipid flip-flip rate decreases in the order of diacylglycerols $>$ cholesterol $>$ other lipids including ceramides. However, amount of data for diacylgycerols (8 simulations with Lipid17 force field) and ceramide (3 simulations with CHARMM force field) is less than for cholesterol (83 simulations), thus we cannot fully exclude the effect of force field or composition on this comparison.

Nevertheless, we use the wide range of available simulations to analyse how cholesterol flip-flop rate depends on membrane properties. Cholesterol flip-flop rates and their averages over fixed ranges of x-axis values are plotted as a function of membrane thickness, lateral density and order in Figs.~\ref{fig:flip-flops} B-D. The results reveal a non-linear correlation between decreasing cholesterol flip-flop rate and membrane packing. Flip-flop rates increase with an order of magnitude when membrane packing density decreases and a major jump is observed at low membrane packing. The order of magnitude changes in cholesterol flip-flop rate with the membrane composition may have major implications in understanding lipid trafficking and membrane biochemistry~\cite{gu19,baral20}. Because the results from the NMRlipids databank are averaged over large range of membrane compositions and force fields, they show that the strong dependence of cholesterol flip-flop rate on membrane properties is not limited to certain lipid compositions or force fields used in previous studies~\cite{gu19,javanainen19,baral20}.

@@

Cholesterol flip-flops are phenomena whose analyses from MD simulations has only recently became possible for groups with substantial resources and special expertise available~\cite{gu19,javanainen19,baral20}. Our analysis demonstrates how the NMRlipids databank makes analyses of such rare phenomena accessible to wide range of scientists in different fields of science and industry who do not have the access to required resources and expertise to produce the large amounts of MD simulation data.

\subsection{Extending the scope of MD simulations to new fields using the NMRlipids databank: Water diffusion anisotropy in membrane systems}
The anisotropy in water diffusion in parallel and perpendicular directions with respect to membranes plays a role in the drug translocation through biological material, particularly in skin \cite{hansen13,wen18,nitsche19,roberts21}, and in MRI imaging~\cite{topgaard20}. MD simulations are rarely used to analyze anistropic diffusion of water as only few permeation events for water are typically observed in a single MD simulation trajectory~\cite{venable19,camilo2022}, thereby making the collection of sufficient amount data challenging. Here we show that the NMRlipids databank can be used to analyze how anisotropic water diffusion in a multilamellar membrane systems depends on membrane properties.

@@

To this end, we first calculated the water permeability through membranes from all simulations in the NMRlipids databank.  The resulting non-zero values range between 0.3 and 322\,$\mu$m/s with the mean and median of 14\,$\mu$m/s and 8\,$\mu$m/s, respectively. These values agree with the previously reported simulation results~\cite{venable19,camilo2022} and have the same order of magnitude as experimentally determined diffusive permeability coefficients, but are on average larger than the values reported for PC lipids in liquid crystalline phase,~0.19-0.33\,$\mu$m/s~\cite{jansen95}. Observed permeabilities and their averages over fixed ranges of values at x-axis are shown in Figs.~\ref{fig:permeability} B-E as a function of temperature, membrane thickness, area per lipid, and acyl chain order.  
%on average, we calculated the average permeabilities over all systems with the value of temperature, thickness, area per lipid, or acyl chain order within a fixed range. These averaged permeabilities also shown in Figs.~\ref{fig:permeability} with black dots. 
As expected, the permeability increases with the temperature, giving 17\,$\pm$\,4\,k$_b$T for the average energy barrier for the water permeation from the Arrhenius plot in Fig.~\ref{fig:permeability} B. On the other hand, the permeability of water decreases on average when membranes become more packed, i.e., with decreasing area per lipid and increasing thickness and acyl chain order (Figs.~\ref{fig:permeability} C-E). Permeation of water through bilayers depends on membrane properties also according previous studies, but it is not agreed whether the area per lipid~\cite{nagle08} or bilayer thickness~\cite{frallicciardi22} would be the main parameter determining the permeability. 
%is previously reported from simulations and experiments for some systems, but not for all~\cite{mathai07,venable19,frallicciardi22}. 
Our analysis over the NMRlipids databank, containing significantly more data than that was available in previous studies, suggest non-linear dependence on both of these parameters, yet the linear correlation would be a good approximation for thicknesses above $\sim$\,3.9\,nm and area per lipids below $\sim$69\,Å$^2$ (insets in Figs.~\ref{fig:permeability} C-D).
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on charged lipids, cholesterol, POPE, or hydration level were not observed in Fig.~\ref{fig:permeationSI} in the supplementary information.
%, although weak decrease for the latter two may be visible.
%
%
%
%%Several models on how water dynamics depends on membrane properties have been proposed based on different experimental and %%theoretical methods, yet consensus 
%%%on how water permeability depends on membrane properties, such as area per lipid and thickness, or molecular composition, 
%%has not been reached~\cite{mathai07,nitsche13,nitsche16,shinoda16,venable19,frallicciardi22}.

To analyze how water diffusion anisotropy depends on membrane properties in a multi-lamellar lipid bilayer system, we calculated also the water diffusion parallel to the membrane surface from all simulations in the NMRlipids databank. The parallel diffusion coefficient of water, $D_{||}$, decreases with reduced hydration and increases with the temperature, but dependencies on membrane area per lipid, thickness, or fraction of charged lipids were not observed in Figs.~\ref{fig:permeability} and~\ref{fig:diffusionSI}.
%The experimental water diffusion increases with increasing hydration level toward the value for bulk water . 
%As expected, water translational diffusion increase in the diffusion is observed with the increased temperature and hydration level (Figs.~\ref{fig:permeability} H and~\ref{fig:diffusionSI} B). 
%The results at different hydration levels are shown in Fig.~\ref{fig:permeability} F together with the experimental data~\cite{rudakova04}. 
Simulation results are close to the experimental values with low hydration levels in Fig.~\ref{fig:permeability} F, but increase to the values of approximately two times higher than experimental bulk water diffusion value (3.1$\cdot$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels. This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{pathirannahalage21}. To estimate the diffusion anisotropy of water, $D_{\perp}/D_{||}$, in multi-lamellar membrane system, the permeability coeffients of water through membranes were translated to perpendicular diffusion coefficients, $D_{\perp}$, using the Tanner equation~\cite{tanner78,wasterby02}. The resulting perpendicular diffusion coefficients are approximately five orders of magnitude smaller than lateral diffusion coefficients of water (Figs.~\ref{fig:permeability} G and H), which is at the upper limit of the anisotropic estimated from the experimental data~\cite{nitsche19}. 
%Large anisotropy values are understandable as simulations give slightly slower permeation rates and higher lateral diffusion rates than experiments. 
Significant increase in the diffusion anisotropy with membrane packing is observed, as $D_{\perp}/D_{||}$ drifts away from one with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability} G and H. This follows from decreasing water permeability with membrane packing (Figs \ref{fig:permeability} C and D), while lateral diffusion remains approximately constant (Fig. \ref{fig:diffusionSI} A and C).

In conclusion, our results suggest that the bilayer packing has a substantial effect on anisotropic water diffusion in multi-membrane lipid systems. Several folds larger anisotropy in membranes with higher lateral density is expected to play a role in pharmagokinetic models not only for water but also for other hydrophilic molecules~\cite{nitsche19}. Furthermore, the understanding of this anisotropy may help in developing new MRI imaging methods~\cite{topgaard20}.

>>>>

%\subsection{Finding the best models for PC and PE lipid mixtures}
The power of the NMRlipids databank in selecting the best simulation parameters for a specific application is demonstrated for mixtures of PE and PC lipids in Fig.~\ref{fig:POPCPOPEapls}, showing areas per lipid from available POPC and POPE bilayer simulations in the databank from different force fields at 310\,K. Predictions from different force fields deviate significantly in terms of absolute values and the slope of decrease upon addition of POPE. Because the area per lipid correlates with the average order parameter of the $\it{sn}$-1 chain (Fig.~\ref{fig:quality} F), simulations with the best predictions for area per lipids can be selected based on order parameter quality evaluation. Quality evaluation in Fig.~\ref{fig:quality} and direct comparison in Fig.~\ref{fig:POPC_POPE_dataSI} reveal that Slipids ranks 1st for POPC with a clear advantage to the others, and 2nd for POPE with only marginally lower quality score than GROMOS-CKP and is therefore the best avilable choice for studies where packing effects of PE lipids are important. 
%The direct comparison with experiments in Fig.~\ref{fig:POPC_POPE_dataSI} in the supplementary information reveals that 
Simulations with CHARMM36 and GROMOS-CKP predict overly packed bilayer for POPC (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}), while the area per lipid for POPC is overstimated in Lipid17. For POPE, 
%order parameters in  Slipids and GROMOS-CKP are close to experiments suggesting that their are per lipids are most reasonable, while
CHARMM36 and Lipid17 predict membranes that are too packed.

In conclusion, 
%the quality evaluation based on the NMRlipids databank suggests that the Slipids parameters would give the most reliable description for membrane packing of POPC:POPE mixtures. Slipids was a relatively good model also for POPC:POPS mixtures in similar comparisons utilizing the preliminary version of the NMRlipids databank~\cite{antila22b}, although charged membranes are more complicated as the counterion binding affinity plays a significant role in membrane packing~\cite{antila19}. On the other hand, CHARMM36 overcomes Slipids when the focus is on conformational ensembles of headgroups  in different types of lipids because the accuracy of glycerol backbone is low in Slipids~\cite{bacle21}. Altogether, these examples demonstrate how 
the automatic quality evaluation in the NMRlipids databank enables rapid exploration and selection of the best simulation parameters for specific applications without extensive and tedious manual force field evaluation. This possibility will promote more reliable simulation results for wide range of applications, support force field development and parametrization of coarse grained force fields against the best atomistic MD simulations. The automatically extracted quantitative quality measures will be particularly useful to guide automatic parametrization procedures.

@@

\subsection{Detection rare phenomena using NMRlipids databank: Cholesterol flip-flops}
Lipid flip-flops from one bilayer leaflet to another play an important role in lipid trafficking and regulating membrane properties~\cite{vanmeer08}. Phospholipid flip-flops are slow when not facilitated by proteins, occuring spontaneously on the timescale of hours or days, while cholesterol, diacylglycerol and ceramide flip-flops are faster. Still, the reported timescales range between minutes to sub-millisecods~\cite{vanmeer08,steck12,parisio16,gu19}. These timescales were previously accessible only by coarse-grained simulations or free energy calculations~\cite{parisio16}, yet atomistic simulations reporting cholesterol flip-flop events have been published recently~\cite{gu19,javanainen19,baral20}. 
%However, systematic studies on how cholesterol flip-flop rate depends on membrane properties are still beyond the current computational resources available for a single research group. 
These studies report an increase in cholesterol flip-flop rates with increasing unsaturation level and decreasing cholesterol concentration~\cite{gu19,javanainen19}, but correlations between cholesterol flip-flop rates and membrane properties have not been systematically studied. Here, we demonstrate that the large amount of simulation data available in the NMRlipids databank with programmatic access makes the analysis of these correlations easily accessible for all.

Flip-flops were observed for cholesterol, % (634 events in 77 different simulations), 
DCHOL (18,19-di-nor-cholesterol), %, 16 flip-flops in 1 simulation) 
DOG (1,2-dioleoyl-sn-glycerol), %, 44 flip-flops in 4 simulations) 
and SDG (1-stearoyl-2-docosahexaenoyl-sn-glycerol) in the NMRlipids databank. Cholesterol flip-flop rates range between 0.001-1.6\,$\mu$s$^{-1}$ with the mean of 0.16\,$\mu$s$^{-1}$ and median of 0.07\,$\mu$s$^{-1}$. These values are consistent with the previously reported values from atomistic MD simulations~\cite{gu19,javanainen19,baral20}. Flip-flop rates of DCHOL, 0.2\,$\mu$s$^{-1}$, was close to the average value of cholesterol, while average rates for diacylglycerols DOG and SDG were higher than for cholesterol, 0.4\,$\mu$s$^{-1}$ and 0.5\,$\mu$s$^{-1}$, respectively. Flip-flops were not observed for other lipids, giving the upper limits for PC lipid flip-flop rate of 9\,s$^{-1}$ and for ceramide (N-palmitoyl-D-erythro-sphingosine) of 0.002\,$\mu$s$^{-1}$. Thus, avilable data in the NMRlipids databank suggest that the lipid flip-flip rate decreases in the order of diacylglycerols $>$ cholesterol $>$ other lipids including ceramides. However, the amount of data for diacylgycerols (8 simulations with Lipid17 force field) and ceramide (3 simulations with CHARMM force field) is less than for cholesterol (83 simulations) so we cannot fully exclude the effect of force field or composition on this comparison.

Nevertheless, we use the wide range of available simulations to analyse how cholesterol flip-flop rate depends on membrane properties. Cholesterol flip-flop rates and their averages over fixed ranges of $x$-axis values are plotted as a function of membrane thickness, lateral density and order in Figs.~\ref{fig:flip-flops} B-D. The results reveal a non-linear correlation between decreasing cholesterol flip-flop rate and membrane packing. Flip-flop rates increase with an order of magnitude when membrane packing density decreases and a major jump is observed at low membrane packing. The orders of magnitude change in cholesterol flip-flop rate with the membrane composition may have major implications in understanding lipid trafficking and membrane biochemistry~\cite{gu19,baral20}. Because the results from the NMRlipids databank are averaged over large range of membrane compositions and force fields, they show that the strong dependence of cholesterol flip-flop rate on membrane properties is not limited to certain lipid compositions or force fields used in previous studies~\cite{gu19,javanainen19,baral20}.

@@

MD simulation and exploration of the cholesterol flip-flops phenomenon has only recently become possible for research groups with substantial resources and specialized expertise~\cite{gu19,javanainen19,baral20}. Our analysis demonstrates how the NMRlipids databank makes analyses of such rare phenomena accessible to wide range of scientists in different fields of science and industry who do not have the access to required resources and expertise to produce the large amounts of MD simulation data required.

\subsection{Extending the scope of MD simulations to new fields using the NMRlipids databank: Water diffusion anisotropy in membrane systems}
The anisotropy in water diffusion in parallel and perpendicular directions with respect to membranes plays a role in the drug translocation through biological material, particularly in skin \cite{hansen13,wen18,nitsche19,roberts21} and in MRI imaging~\cite{topgaard20}. MD simulations are rarely used to analyze anistropic diffusion of water since only few permeation events for water are typically observed in a single MD simulation trajectory~\cite{venable19,camilo2022}, making the collection of sufficient data challenging or impossible. Here, we show that the NMRlipids databank can be used to systematically analyze how anisotropic water diffusion in a multilamellar membrane systems depends on membrane properties.

@@

To this end, we calculated the water permeability through membranes from all simulations in the NMRlipids databank. The resulting non-zero values range between 0.3 and 322\,$\mu$m/s with a mean and median of 14\,$\mu$m/s and 8\,$\mu$m/s, respectively. These values agree with the previously reported simulation results~\cite{venable19,camilo2022} and have the same order of magnitude as experimentally determined diffusive permeability coefficients, but are on average larger than the values reported for PC lipids in the liquid crystalline phase,~0.19-0.33\,$\mu$m/s~\cite{jansen95}. Observed permeabilities and their averages over fixed ranges of values at $x$-axis are shown in Figs.~\ref{fig:permeability} B-E as a function of temperature, membrane thickness, area per lipid, and acyl chain order.  
%on average, we calculated the average permeabilities over all systems with the value of temperature, thickness, area per lipid, or acyl chain order within a fixed range. These averaged permeabilities also shown in Figs.~\ref{fig:permeability} with black dots. 
As expected, the permeability increases with the temperature, giving an average energy barrier of 17\,$\pm$\,4\,k$_b$T for the water permeation from the Arrhenius plot in Fig.~\ref{fig:permeability} B. On the other hand, the permeability of water decreases on average when membranes become more packed, i.e., with decreasing area per lipid and increasing thickness and acyl chain order (Figs.~\ref{fig:permeability} C-E). Permeation of water through bilayers also depends on membrane properties according previous studies, but there is no established consensus on whether the area per lipid~\cite{nagle08} or bilayer thickness~\cite{frallicciardi22} is the main parameter determining the permeability. 
%is previously reported from simulations and experiments for some systems, but not for all~\cite{mathai07,venable19,frallicciardi22}. 
Our analysis over the NMRlipids databank, containing significantly more data than that was available in previous studies, suggest non-linear dependence on both of these parameters, yet the linear correlation would be a good approximation for thicknesses above $\sim$\,3.9\,nm and area per lipids below $\sim$69\,Å$^2$ (insets in Figs.~\ref{fig:permeability} C-D).
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on charged lipids, cholesterol, POPE, or hydration level were not observed in Fig.~\ref{fig:permeationSI} in the supplementary information.
%, although weak decrease for the latter two may be visible.
%
%
%
%%Several models on how water dynamics depends on membrane properties have been proposed based on different experimental and %%theoretical methods, yet consensus 
%%%on how water permeability depends on membrane properties, such as area per lipid and thickness, or molecular composition, 
%%has not been reached~\cite{mathai07,nitsche13,nitsche16,shinoda16,venable19,frallicciardi22}.

To analyze how water diffusion anisotropy depends on membrane properties in a multi-lamellar lipid bilayer system, we calculated the water diffusion parallel to the membrane surface from all simulations in the NMRlipids databank. The parallel diffusion coefficient of water, $D_{||}$, decreases with reduced hydration and increases with the temperature, but dependencies on membrane area per lipid, thickness, or fraction of charged lipids were not observed in Figs.~\ref{fig:permeability} and~\ref{fig:diffusionSI}.
%The experimental water diffusion increases with increasing hydration level toward the value for bulk water . 
%As expected, water translational diffusion increase in the diffusion is observed with the increased temperature and hydration level (Figs.~\ref{fig:permeability} H and~\ref{fig:diffusionSI} B). 
%The results at different hydration levels are shown in Fig.~\ref{fig:permeability} F together with the experimental data~\cite{rudakova04}. 
Simulation results are close to the experimental values with low hydration levels in Fig.~\ref{fig:permeability} F, but increase to approximately two times higher than the experimental value for bulk water diffusion (3.1$\cdot$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels. This is not surprising as the most common water model used in membrane simulations, TIP3P, is known to overestimate the bulk water diffusion~\cite{pathirannahalage21}. To estimate the diffusion anisotropy of water, $D_{\perp}/D_{||}$, in multi-lamellar membrane system, the permeability coeffients of water through membranes were translated to perpendicular diffusion coefficients, $D_{\perp}$, using the Tanner equation~\cite{tanner78,wasterby02}. The resulting perpendicular diffusion coefficients are approximately five orders of magnitude smaller than lateral diffusion coefficients of water (Figs.~\ref{fig:permeability} G and H), which is at the upper limit of the anisotropic diffusion as determined by experiments~\cite{nitsche19}. 
%Large anisotropy values are understandable as simulations give slightly slower permeation rates and higher lateral diffusion rates than experiments. 
Significant increase in the diffusion anisotropy with membrane packing is observed, as $D_{\perp}/D_{||}$ drifts away from one with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability} G and H. This follows from decreasing water permeability with membrane packing (Figs \ref{fig:permeability} C and D), while lateral diffusion remains approximately constant (Fig. \ref{fig:diffusionSI} A and C).

In conclusion, our results suggest that the bilayer packing has a substantial effect on anisotropic water diffusion in multi-membrane lipid systems. Several folds larger anisotropy in membranes with higher lateral density is expected to play a role in pharmagokinetic models, not only for water but also for other hydrophilic molecules~\cite{nitsche19}. Furthermore, the enhanced understanding of this anisotropy may help in developing new MRI imaging methods~\cite{topgaard20}.
